---
title: "Cyclitics Analysis"
author: "Jordan Prince"
date: "`r Sys.Date()`"
output: html_document
---

# Cyclitics User Data Analysis

### STEP 1: R Environment Setup
```{r R setup, ehco=TRUE}
# helps to wrangle data
library(tidyverse)
# managing conflicted packages
library(conflicted)
# setting dplyr::filter and dplyr::lag as defaults
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

### STEP 2: Collect Data
```{r data collection, echo=TRUE}
# Upload Divvy data sets (csv files)
q1_2019 <- read_csv("C:/Users/Jordan/Documents/RStudio/Cyclitics/Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("C:/Users/Jordan/Documents/RStudio/Cyclitics/Divvy_Trips_2020_Q1.csv")
```

### STEP 3: Consilidate Into A Single File
```{r consilidation, echo=TRUE}
# Compare column names of each of the files
colnames(q1_2019)
colnames(q1_2020)

# Renaming columns so they are consistent with Q1_2020
(q1_2019 <- rename(q1_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,start_station_name = from_station_name
                   ,start_station_id = from_station_id
                   ,end_station_name = to_station_name
                   ,end_station_id = to_station_id
                   ,member_casual = usertype
                   ))

# Inspecting data frames for any other inconsistencies
str(q1_2019)
str(q1_2020)

# Converting ride_id and rideable_type to character
q1_2019 <- mutate(q1_2019, ride_id = as.character(ride_id)
                  ,rideable_type = as.character(rideable_type))

# Stack individual quarter's frames into one
all_trips <- bind_rows(q1_2019, q1_2020)

# Remove lat, long, birthyear, and gender fields
all_trips <- all_trips %>%
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "tripduration"))
```

### STEP 4: Clean & Add Data To Prepare For Analysis
```{r cleaning, echo=TRUE}
# Inspect the new table
# list of column names
colnames(all_trips)
# number of rows
nrow(all_trips)
# dimensions of data frame
dim(all_trips)
# preview first 6 rows
head(all_trips)
# list of columns and data types
str(all_trips)
# statistical summary of data
summary(all_trips)

# consolidating member_casual values from four to two
# check how many distinct values are in the column
table(all_trips$member_casual)
# rename "Subscriber" and "Customer" to "member" and "casual" respectively
all_trips <- all_trips %>%
  mutate(member_casual = if_else(member_casual == "Subscriber", "member", member_casual))
all_trips <- all_trips %>%
  mutate(member_casual = if_else(member_casual == "Customer", "casual", member_casual))
# Recheck distinct values
table(all_trips$member_casual)

# Add columns: date, month, day, and year
all_trips$date <-as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

# Add a "ride_length" calculation to all_trips
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

# Inspect the structure of the columns
str(all_trips)

# Convert "ride_length" to a time format
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

# Remove bad data
# Creating new version of data because of removal of some data
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```

### STEP 5: Analysis
```{r analysis, echo=TRUE}
# Descriptive analysis on ride_length
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
max(all_trips_v2$ride_length)
min(all_trips_v2$ride_length)
summary(all_trips_v2)

# Compare member and casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

# Avg ride time by each day (members v casuals)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
# Ordering the days of the week
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Rerun avg ride time
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

# Analyze ridership data by type and weekday
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)

# Visualize number of rides by rider type
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

#### Further Analysis
```{r more analysis, echo=TRUE}
# overall number of trips
# scales package makes sure that the plot does not use scientific notation
library(scales)
ggplot(all_trips_v2, aes(x = day_of_week, fill = member_casual)) +
  geom_bar(position = "dodge") +
  labs(x = "Day of Week", y = "Number of Trips", title = "Number of Trips by Members and Casual Riders") +
  scale_y_continuous(labels = comma) +
  theme_minimal()


# Create heat map
library(ggplot2)
library(dplyr)
library(lubridate)

# Create a new column 'hour'
all_trips_v2 <- all_trips_v2 %>%
  mutate(hour = hour(started_at))

# Create a summary table
summary_data <- all_trips_v2 %>%
  group_by(day_of_week, hour, member_casual) %>%
  summarize(trips = n(), .groups = "drop")

# Create line plot with facet grid
ggplot(summary_data, aes(x = hour, y = trips, color = member_casual)) +
  geom_line() +
  facet_grid(day_of_week ~ .) +
  labs(x = "Hour of the Day", y = "Number of Trips", color = "Rider Type", 
       title = "Hourly Trips by Members and Casual Riders") +
  theme_minimal()
```

#### Notes:
##### As you can see, the first bar chart showing the average duration of trips by members and casual riders, it would appear the casual riders are much heavier users of the service than members are. However, when we look at the overall number of trips by members and casual riders we see that members dwarf casual rider numbers significantly. Considering the spike that we see in the number of trips on Saturday and Sunday, along with the data shown in the line chart *Hourly Trips by Members and Casual Riders* we can infer that members use the service to commute to and from work during the weekdays while casual riders more so use the service on the weekends to visit local tourist attractions.

##### The most used stations by casual riders are the following: 
- Streeter Dr & Grand Ave: A Navy Pier with many restaurants, gift shops, and monuments
- Shedd Aquarium
- DuSable Harbor: Which hosts a yacht club
- Buckingham Fountain
- Alder Planetarium
##### All of the above mentioned stations used by casual riders are highly frequented local attractions that provide some sort of leisurely activities such as site-seeing, shopping, boating and more. 

#### RECOMMENDATIONS:
##### 1) First recommendation based on the data insights shown above is to partner with local tourist attractions to give members discounts on goods and services. This would be a great incentive not just for casual riders to become members but also a way to bring more business to the area which in turn would bring the Cyclitics company more business as well.

##### 2) My second recommendation is to push ads in the app and on the company's website with quick and easy to read offers that detail the benefits of being a member. The company may also want to consider limited time but significant discounts for new members, possibly also referral discounts for casual riders that convert to members and remain members for a period of time no less than 30 days or a certain amount of trips.

##### 3) Last, I recommend the installation of large print adverts specifically at the above mentioned stations that are frequented by the casual rides.